The markings of the trade: exploring the patterns of olive oil production in Roman Baetica

### 2020-07-06 Paper/jaccardbaetica.tex
add new changes xavi

The structure and processes associated to this massive olive oil production has been extensively discussed over the last decades\citep{rodriguez_economioleicola_1977, Chic_hispania_1997,millet_anforas_1998}. Our knowledge of this economic activity has benefited from new findings and data sources, but despite all these advances several questions remain open: \xavi{no entiendo esta pregunta} what patterns were followed for the distribution of olive oil to the different provinces? \xavi{no es esta pregunta idéntica a la anterior?} Did each province follow a different pattern for its distribution? \xavi{esta frase no tiene sentido aquí; o la pones antes o en otro párrafo, pero es raro acabar un set de 2 preguntas con una frase que no sea otra pregunta} Nor the lack of written records has made possible to detect any indication of patterns in the olive oil market.

\xavi{las preguntas no van sobre patrones; los patrones son las trazas en la evidencia que te ayudan a responder a las preguntas}

@@

These stamps are typically found on handles but they were also imprinted on rims and the amphora body~\citep{millet_anforas_1998}. The content information of the stamps is shown in different forms and letter content and it seems that there was not a unique criterion
%\xavi{qué information? El código de letras?}. 
Stamps often displayed a code of three letters and they can appear in abbreviated form or complete and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}.\xavi{que quieres decir con abbreviated o complete? Como abrevias un código de 3 letras?}

Scholars agree that stamps are some type of identification mark, but there is no consensus on the meaning of these marks\citep{rodriguez_baetican_1998}. Additionally, a large percentage of Dressel 20 containers were not stamped thus complicating the interpretation~\xavi{esta última frase aquí es un poco random no?}. The stamp codes are interpreted based on three main ideas: content (olive oil), context (amphora workshop) and subject (individuals involved on the production). On the one hand, it seems that stamps could have been identified as the landowner of the olive groves \citep{rodriguez_economioleicola_1977}. On the other hand, they could also belong to the owner of the workshops were the amphorae was made or even to a group of amphora workers~\citep{berni_millet_epigrafianforica_2008}. In any case, the use of these stamps is a good proxy to explore the system of olive oil production and trade developed in Baetica.

@@

The output was a dataset of 4271 stamps found in sites belonging to \textit{Britannia} (2219 stamps) and \textit{Germania} (2052 stamps). Both \textit{Germania} and \textit{Britannia} were analysed as borders and not as Roman provinces. This means that we included in the sample some centres that are actually located beyond the administrative boundaries but were considered part of the same border. Specifically, for the German \textit{limes} we included sites from the two provinces (Superior and Inferior) while the analysis of Britannia extended to \textit{Caledonia}.

%We analysed a dataset of 2219 stamps from different centres in \textit{Britannia}. 
%CITAR (Callender, 1965; Carreras Monfort y Funari,1998; Ayllón-Martı́ et al., 2018). 
In \textit{Britannia}, we studied a total sample of 1765 stamps from 46 sites displaying 968 unique codes\xavi{pero esto no concuerda con las 2219 de antes...es por descartar los inciertos? Hay que dejarlo claro}. The centres located in \textit{Britannia} can be seen in Figure~\ref{britannia}.

@@

@@

The results for \textit{Germania} follow a similar pattern than \textit{Britannia} but with a minor correlation as we can observe in the dendrogram\xavi{esta frase está repetida de los resultados. quizá quitarla de ahí y dejarla aquí?}. The areas mostly militarised share similar stamps than civil areas. However, we do not detect a concrete pattern regarding the distribution of the stamps in the German limes \citep{xanten2018}.

@@

To conclude, the method presented here provides a useful framework to improve our understanding of the organisation and processes that allowed the Roman Empire to manage a massive and highly specialized production of essential goods.  This work has identified differences in the case of the amphoric production within the Roman Empire\xavi{qué diferencias?}. The growing amount of archaeological data requires of large-scale quantitative methods such as the one presented here to identify and explore complex patterns, thus allowing us to infer dynamics and improve our interpretation of complex economic processes of past societies.

>>>>

The structure and processes associated to this massive olive oil production has been extensively discussed over the last decades\citep{rodriguez_economioleicola_1977, Chic_hispania_1997,millet_anforas_1998}. 
However, the lack of written records has made possible to detect any indication of patterns in the olive oil market.

Our knowledge of this economic activity has benefited from new findings and data sources, but despite all these advances several questions remain open: How was amphora production organized during the Roman Empire? Can we detected economic dynamics between provinces by analysing the distribution of amphorae?

%Did each province follow a different pattern for its distribution? 
%\xavi{no entiendo esta pregunta} what patterns were followed for the distribution of olive oil to the different provinces? 
%\xavi{no es esta pregunta idéntica a la anterior?} 
%\xavi{esta frase no tiene sentido aquí; o la pones antes o en otro párrafo, pero es raro acabar un set de 2 preguntas con una frase que no sea otra pregunta}

%\xavi{las preguntas no van sobre patrones; los patrones son las trazas en la evidencia que te ayudan a responder a las preguntas}

@@

These stamps are typically found on handles but they were also imprinted on rims and the amphora body~\citep{millet_anforas_1998}. The content information of the stamps is shown in different forms and letter content and it seems that there was not a unique criterion.
%\xavi{qué information? El código de letras?}. 
Stamps often displayed a code of three letters and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}. The three letters correspond to a initial forms of a name but they can also appear complete. 
%\xavi{que quieres decir con abbreviated o complete? Como abrevias un código de 3 letras?}

Scholars agree that stamps are some type of identification mark, but there is no consensus on the meaning of these marks\citep{rodriguez_baetican_1998}.

%Additionally, a large percentage of Dressel 20 containers were not stamped thus complicating the interpretation~\xavi{esta última frase aquí es un poco random no?}.

The stamp codes are interpreted based on three main ideas: content (olive oil), context (amphora workshop) and subject (individuals involved on the production). On the one hand, it seems that stamps could have been identified as the landowner of the olive groves \citep{rodriguez_economioleicola_1977}. On the other hand, they could also belong to the owner of the workshops were the amphorae was made or even to a group of amphora workers~\citep{berni_millet_epigrafianforica_2008}. In any case, the use of these stamps is a good proxy to explore the system of olive oil production and trade developed in Baetica.

@@

The output was a dataset of 4271 stamps found in sites belonging to \textit{Britannia} (total: 2219 stamps, analysed: 1765 stamps ) and \textit{Germania} (total: 2052 stamps, analysed: 1621 stamps). Some stamps from \textit{Britannia} and \textit{Germania} were discarded for the analysis for incomplete information.

Both \textit{Germania} and \textit{Britannia} were analysed as borders and not as Roman provinces. This means that we included in the sample some centres that are actually located beyond the administrative boundaries but were considered part of the same border. Specifically, for the German \textit{limes} we included sites from the two provinces (Superior and Inferior) while the analysis of Britannia extended to \textit{Caledonia}.

%We analysed a dataset of 2219 stamps from different centres in \textit{Britannia}. 
%CITAR (Callender, 1965; Carreras Monfort y Funari,1998; Ayllón-Martı́ et al., 2018). 
In \textit{Britannia}, we studied a total sample of 1765 stamps from 46 sites displaying 968 unique codes

%\xavi{pero esto no concuerda con las 2219 de antes...es por descartar los inciertos? Hay que dejarlo claro}.

The centres located in \textit{Britannia} can be seen in Figure~\ref{britannia}.

@@

The results for \textit{Germania} follow a similar pattern than \textit{Britannia} but with a minor correlation as we can observe in the dendrogram.

@@

%The results for \textit{Germania} follow a similar pattern than \textit{Britannia} but with a minor correlation as we can observe in the dendrogram\xavi{esta frase está repetida de los resultados. quizá quitarla de ahí y dejarla aquí?}.

.The areas mostly militarised share similar stamps than civil areas. However, we do not detect a concrete pattern regarding the distribution of the stamps in the German limes \citep{xanten2018}.

@@

To conclude, the method presented here provides a useful framework to improve our understanding of the organisation and processes that allowed the Roman Empire to manage a massive and highly specialized production of essential goods. This work has identified similarities between amphoric stamps and spatial distribution in the case of the amphoric production within the Roman Empire
%\xavi{qué diferencias?}. 
The growing amount of archaeological data requires of large-scale quantitative methods such as the one presented here to identify and explore complex patterns, thus allowing us to infer dynamics and improve our interpretation of complex economic processes of past societies.
